# Basic Reactivity

```{r setup , message=FALSE, echo=FALSE}
library(DiagrammeR)
```

## Recap

```{r intro-graph, echo=FALSE}
mermaid("
graph LR
A[R scripting]-->B[sequential logic]
")

mermaid("
graph LR
A[Reactive Programming]-->B[graph of dependencies]
")
```

Important Learnings so far, main app components:

1. Front end - the ui object

-   contains the HTML presented to every user of your app.
-   is simple because **every user** gets the **same** HTML.

2.  Back end - the server object

-   is more complicated because **every user** needs to get an **independent version** of the app (when user A modifies an input field, user B shouldn't see their outputs change).
-   creates a new environment for each run, giving each session to have a unique state.      


> server <- function(input, output, session)

1 _input_       
- list-like object       
- used for receiving input( sent from the browser)       
- read-only       
- must be read in a reactive context (e.g. renderText() or reactive())      


2 _output_     
- list-like object      
- used for sending output       
- ALWAYS use with a render fn() - sets up the reactive context & renders the HTML.      


```{r eval=FALSE}
ui <- fluidPage(
  textInput("name", "What's your name?"),
  textOutput("greeting")
)

server <- function(input, output, session) {
  output$greeting <- renderText({
    paste0("Hello ", input$name, "!")
  })
}
```

## Reactive Programming

Mental Model       
- tell vs inform (providing Shiny with recipes, not giving it commands).        

### Imperative vs declarative programming

This difference between commands and recipes is one of the key differences between two important styles of programming:         
- _Imperative_ programming - issue a specific command and it’s carried out immediately.       
- _Declarative_ programming - express higher-level goals or describe important constraints, and rely on someone else to decide how and/or when to translate that into action.       

- Imperative code is assertive; vs declarative code is passive-aggressive       
- “Make me a sandwich” vs “Ensure there is a sandwich in the refrigerator whenever I look inside of it”       

In essesnce, you describe your overall goals, and the software figures out how to achieve them without further intervention. 

### Laziness

It allows apps to be extremely lazy. A Shiny app will only ever do the minimal amount of work needed to update the output controls that you can currently see.

_CAUTION_: If you’re working on a Shiny app and you just can’t figure out why your code never gets run, double check that your UI and server functions are using the same identifiers. 


### The reactive graph

- understanding order of execution
- code is only run when needed

_reactive graph_ 
- describes how inputs and outputs are connected
- describe this relationship (`output` has a reactive dependency on `input`)


```{r reactive-graph}
mermaid("
graph LR
A[name] --- B>greeting]
")
```

reactive graph is a powerful tool for understanding how your app works.       
- Make by hand     
- Use DiagrammeR pkg to make it manually yourself     
- Use `reactlog` pkg to do it automatically (more in later chapters)       

## Reactive expressions

**What** - A tool that **reduces duplication** in your reactive code by introducing additional nodes into the reactive graph       
**How** - using `reactive()`      


```{r}
# Just for example

server <- function(input, output, session) {
  string <- reactive(paste0("Hello ", input$name, "!"))
  output$greeting <- renderText(string())
}
```

```{r reactive-graph-with-reactive() , echo=FALSE}
mermaid("
graph LR
A[name] --- B>string]
B>string] --- C>greeting]
")
```

In other words , makes app cleaner & more efficient (by removing redundant codes & recomputation).
It also simplifies the reactive graph.

Reactive expressions have a flavour of both inputs and outputs:

- Like inputs, you can use the results of a reactive expression in an output.
- Like outputs, reactive expressions depend on inputs and automatically know when they need updating.

New vocab:

- producers to refer to reactive inputs and expressions, and 
- consumers to refer to reactive expressions and outputs


**Execution order**         
- determined solely by the reactive graph (and not the order of lines of code/layout in the _server_ fn unlike normal R scripts)


## Controlling timing of evaluation

**_Timed invalidation_**        
**How** - using `reactiveTimer()`


**_On click_**       
**How** - using `actionButton()`, `eventReactive()`      

## Observers

There are two important differences between `observeEvent()` and `eventReactive()`:

1.  You don't/can't assign the result of observeEvent() to a variable, so\
2.  You can't refer to it from other reactive consumers.

## Execises

```{r example}
mermaid("
graph LR
A(Rounded)-->B[Rectangular]
B-->C{A Rhombus}
C-->D[Rectangle One]
C-->E[Rectangle Two]
")
```

```{r reactive-graph-exercise}
mermaid("
graph LR
A[A arr_text] --- B>A arrowtext]
")
```
