# Layout, themes, HTML

```{r}
library(htmltools)
library(shiny)
```

## Disclaimer

"Mastering Shiny" is a moving target.

Much of the content for "Layout, themes, HTML" chapter was originally part of
the "Basic UI" chapter.

<!-- So I'm attempting to put a web-dev oriented spin on the material -->

Hence, these notes have greater emphasis on low-level web concerns (html / css /
bootstrap)

## Learning objectives

- Creating raw html using R
- HTML elements: attributes, classes, content
- CSS for styling
- Bootstrap (and related) front end toolkits
- Shiny produces single-page applications
- ... but multipage _layouts_ are possible

## Resources

### Websites

- [Awesome Shiny Extensions](https://github.com/nanxstats/awesome-shiny-extensions)
- [Shiny Application Layout Guide](https://shiny.rstudio.com/articles/layout-guide.html)

- Mozilla Developer Network
  - [Intro to HTML](https://developer.mozilla.org/en-US/docs/Learn/HTML/Introduction_to_HTML)
  - [CSS First Steps](https://developer.mozilla.org/en-US/docs/Learn/CSS/First_steps)
  - [Website parsing figure](https://developer.mozilla.org/en-US/docs/Learn/CSS/First_steps/How_CSS_works/rendering.svg)

### Books

- [Outstanding User Interfaces ...](https://unleash-shiny.rinterface.com)
  - Especially Chapters 1, 5, 6, 7

### Packages / Tools

- [Bootstrap](https://getbootstrap.com/)
- [Sass](https://sass-lang.com/)

- R packages
  - `{htmltools}` (for building raw html)
  - `{sass}` (for building css using Sass rules; not mentioned in Wickham)
  - `{bslib}` (handles working with bootstrap)
  - `{thematic}` (matches plotting theme to app theme)
  - `{shiny.semantic}` (formantic)
  - `{shinyMobile}` (framework 7)
  - `{shinymaterial}` (Material design)
  - `{shinydashboard}` (for building dashboards)

## Under the hood

A typical web app:

- In the browser (front-end):
  - .html file defines the content
  - .css files define the style
  - javascript handles interactivity

- On the server (back-end):
  - Requests received from the client
  - R performs computations based on those requests
  - Sends responses to the client

### HTML - Hypertext markup language

To view the .html for an app

- Open an app in the browser or Rstudio
- "View source" / "Inspect" / Open "Developer Tools"

Typical .html structure:

```html
<!DOCTYPE html>
<html>
  <head>
    <!-- Metadata goes here (encodings, dependencies, author ...) -->
    <meta charset="utf-8">
    <title>The page title</title>
  </head>
  <body>
    <!-- Content goes here!-->
    <h1>My Printed Title</h1>
    <p>
      A paragraph of text.
    </p>
  </body>
</html>
```

### Writing HTML with `shiny` / `htmltools`

How to write .html using just R code?

The simplest shiny UI:

```{r}
ui <- shiny::fluidPage()
```

The .html body for that UI can be viewed directly:

```{r, comment = ""}
cat(
  as.character(ui)
)
```

To also include the 'head' for the .html:

```{r, comment=""}
cat(
  shiny:::renderPage(ui)
)
```

`script` and `link` elements in the `<head>` = dependencies for the site

Adding bare HTML to the shiny-produced HTML

```{r, comment=""}
ui_with_raw_html <- shiny::fluidPage(
  # some raw html
  htmltools::HTML(r"(
    <h1>This is a heading</h1>
    <ul>
      <li>First bullet</li>
      <li>Second bullet</li>
    </ul>
    )"
  ),
  # a shiny text imput
  textInput("name", "What's your name?")
)

cat(
  as.character(ui_with_raw_html)
)
```

Using HTML helpers:

```{r}
ui_from_html_helpers <- shiny::fluidPage(
  h1("This is a heading", class = "my-class"),
  tags$ul(
    tags$li("First bullet"),
    tags$li("Second bullet")
  ),
  textInput("name", "What's your name?")
)

cat(
  as.character(ui_from_html_helpers)
)
```

### CSS - Cascading Style Sheets

CSS

- isn't covered in the chapter (but links provided)
- but underlies how bootstrap styles your shiny app

Much more detail at

- [MDN](https://developer.mozilla.org/en-US/docs/Learn/CSS);
- and in [Outstanding User Interfaces with Shiny](https://unleash-shiny.rinterface.com/beautify-css.html)

How does the browser display a site / app?

[MDN Website parsing figure](https://developer.mozilla.org/en-US/docs/Learn/CSS/First_steps/How_CSS_works/rendering.svg)

### Using CSS to style a shiny app

A simple CSS file to set any text with "my-class" as class to red

```css
/* ./examples/06-layout_themes_html-css_example/www/style.css */
.my-class {
  color: red
}
```

- Put static files (javascript / css) in the `./www` subdirectory of your app
- Include your files using `tags$link` when defining the UI
- Note: the path in the link should be "style.css" not "www/style.css"

```r
# ./examples/06-layout_themes_html-css_example/app.R
library(shiny)

ui <- shiny::fluidPage(
  tags$link(rel = "stylesheet", type = "text/css", href = "style.css"),
  h1("This is a heading", class = "my-class"),
  textInput("name", "What's your name?")
)
server <- function(input, output, session) {}

shinyApp(ui, server)
```

You can write custom rules to style based on

- html element (`p`, `h1`, ...),
- attributes
- IDs 
- and on interactive things (eg, when the mouse hovers)

## Bootstrap & Themes

<!--

## Book Structure

Book sections:
- Single page layouts (+Ex; this section was previously in the UI chapter, and
  was covered in an earlier bookclub week)
- Multi page layouts
- Bootstrap
- Themes (+Ex)
- Under the hood

? When presenting, maybe reverse the order of the sections:

- Learning Outcomes
- Under the hood
  - HTML
  - CSS
- Bootstrap & Themes
- Layouts
  - Single page
  - Multi page

TODO: think up an exercise that uses multi-page layouts
-->
