# Layout, themes, HTML

```{r}
library(bslib)
library(htmltools)
library(shiny)
```

## Disclaimer

"Mastering Shiny" is a moving target.

Much of the content for "Layout, themes, HTML" chapter was originally part of
the "Basic UI" chapter.

<!-- So I'm attempting to put a web-dev oriented spin on the material -->

Hence, these notes have greater emphasis on low-level web concerns (html / css /
bootstrap)

## Learning objectives

- Creating raw html using R
- HTML elements: attributes, classes, content
- CSS for styling
- Bootstrap (and related) front end toolkits
- Shiny produces single-page applications
- ... but multipage _layouts_ are possible

## Resources

### Websites

- [Awesome Shiny Extensions](https://github.com/nanxstats/awesome-shiny-extensions)
- [Shiny Application Layout Guide](https://shiny.rstudio.com/articles/layout-guide.html)

- Mozilla Developer Network
  - [Intro to HTML](https://developer.mozilla.org/en-US/docs/Learn/HTML/Introduction_to_HTML)
  - [CSS First Steps](https://developer.mozilla.org/en-US/docs/Learn/CSS/First_steps)
  - [Website parsing figure](https://developer.mozilla.org/en-US/docs/Learn/CSS/First_steps/How_CSS_works/rendering.svg)

### Books

- [Outstanding User Interfaces ...](https://unleash-shiny.rinterface.com)
  - Especially Chapters 1, 5, 6, 7

### Packages / Tools

- [Bootstrap](https://getbootstrap.com/)
- [Sass](https://sass-lang.com/)

- R packages
  - `{htmltools}` (for building raw html)
  - `{sass}` (for building css using Sass rules; not mentioned in Wickham)
  - `{bslib}` (handles working with bootstrap)
  - `{thematic}` (matches plotting theme to app theme)
  - `{shiny.semantic}` (formantic)
  - `{shinyMobile}` (framework 7)
  - `{shinymaterial}` (Material design)
  - `{shinydashboard}` (for building dashboards)

## Under the hood

A typical web app:

- In the browser (front-end):
  - .html file defines the content
  - .css files define the style
  - javascript handles interactivity

- On the server (back-end):
  - Requests received from the client
  - R performs computations based on those requests
  - Sends responses to the client

### HTML - Hypertext markup language

To view the .html for an app

- Open an app in the browser or Rstudio
- "View source" / "Inspect" / Open "Developer Tools"

Typical .html structure:

```html
<!DOCTYPE html>
<html>
  <head>
    <!-- Metadata goes here (encodings, dependencies, author ...) -->
    <meta charset="utf-8">
    <title>The page title</title>
  </head>
  <body>
    <!-- Content goes here!-->
    <h1>My Printed Title</h1>
    <p>
      A paragraph of text.
    </p>
  </body>
</html>
```

### Writing HTML with `shiny` / `htmltools`

How to write .html using just R code?

The simplest shiny UI:

```{r}
ui <- shiny::fluidPage()
```

The .html body for that UI can be viewed directly:

```{r, comment = ""}
cat(
  as.character(ui)
)
```

To also include the 'head' for the .html:

```{r, comment=""}
cat(
  shiny:::renderPage(ui)
)
```

`script` and `link` elements in the `<head>` = dependencies for the site

Adding bare HTML to the shiny-produced HTML

```{r, comment=""}
ui_with_raw_html <- shiny::fluidPage(
  # some raw html
  htmltools::HTML(r"(
    <h1>This is a heading</h1>
    <ul>
      <li>First bullet</li>
      <li>Second bullet</li>
    </ul>
    )"
  ),
  # a shiny text imput
  textInput("name", "What's your name?")
)

cat(
  as.character(ui_with_raw_html)
)
```

Using HTML helpers:

```{r}
ui_from_html_helpers <- shiny::fluidPage(
  h1("This is a heading", class = "my-class"),
  tags$ul(
    tags$li("First bullet"),
    tags$li("Second bullet")
  ),
  textInput("name", "What's your name?")
)

cat(
  as.character(ui_from_html_helpers)
)
```

### CSS - Cascading Style Sheets

CSS

- isn't covered in the chapter (but links provided)
- but underlies how bootstrap styles your shiny app

Much more detail at

- [MDN](https://developer.mozilla.org/en-US/docs/Learn/CSS);
- and in [Outstanding User Interfaces with Shiny](https://unleash-shiny.rinterface.com/beautify-css.html)

How does the browser display a site / app?

[MDN Website parsing figure](https://developer.mozilla.org/en-US/docs/Learn/CSS/First_steps/How_CSS_works/rendering.svg)

### Using CSS to style a shiny app

A simple CSS file to set any text with "my-class" as class to red

```css
/* ./examples/06-layout_themes_html-css_example/www/style.css */
.my-class {
  color: red
}
```

- Put static files (javascript / css) in the `./www` subdirectory of your app
- Include your files using `tags$link` when defining the UI
- Note: the path in the link should be "style.css" not "www/style.css"

```r
# ./examples/06-layout_themes_html-css_example/app.R
library(shiny)

ui <- shiny::fluidPage(
  tags$link(rel = "stylesheet", type = "text/css", href = "style.css"),
  h1("This is a heading", class = "my-class"),
  textInput("name", "What's your name?")
)
server <- function(input, output, session) {}

shinyApp(ui, server)
```

You can write custom rules to style based on

- html element (`p`, `h1`, ...),
- attributes
- IDs 
- and on interactive things (eg, when the mouse hovers)

## Bootstrap & Themes

Shiny uses the [bootstrap](https://getbootstrap.com/) toolkit for styling.

Here's the strapline for bootstrap:

    Quickly design and customize responsive mobile-first sites with Bootstrap,
    the world’s most popular front-end open source toolkit, featuring Sass
    variables and mixins, responsive grid system, extensive prebuilt components,
    and powerful JavaScript plugins.

What does that mean?

- Responsive
  - The page alters layout / appearance for different screen widths
- Sass variables and mixins
  - These are tools that reduce the amount of repetition when writing CSS
- Prebuilt components
  - You don't have to write your own button, navbar, etc ... classes from
  scratch
- Javascript plugins
  - eg, for using dropdowns

So bootstrap solves lots of challenges for us

Use package `{bslib}` to work with bootstrap in R

To modify themes:

```r
bslib::bs_theme(
  # colours in RGB
  bg = "#202123", # background
  fg = "#B8BCC2", # foreground
  primary = NULL,
  secondary = NULL,
  success = NULL,
  info = NULL,
  warning = NULL,
  danger = NULL,
  # fonts
  base_font = list(bslib::font_google("Pacifico", local = TRUE), "sans-serif"),
  code_font = NULL,
  heading_font = NULL
  # further options ...
)
```

Or use `bootswatch` to pick a prebuilt theme:

```r
bslib::bs_theme(bootswatch = "sandstone")
```

Then use a theme in your UI definition:

```{r, comment = ""}
theme_dark <- bslib::bs_theme(
  bootswatch = "darkly",
  base_font = list(bslib::font_google("Pacifico", local = TRUE), "sans-serif")
)

ui_with_bs <- fluidPage(
  theme = theme_dark
)

cat(
  shiny:::renderPage(ui_with_bs)
)
```

## Layouts

The Layouts section was moved from the Basic UI chapter with little change.

Rather than go over them again, we'll rewrite a simple (single-page layout) app
into something more complicated (and style it while we're at it)

See `./examples/06-layout_themes_html-single_to_multipage_app/`

### Single-page Layouts

Page functions: `fluidPage`, `fixedPage`, `fillPage`

Layout functions: `sidebarLayout`, `fluidRow`, `column`

### Multi-page Layouts

Page functions: `navbarPage` (or nest the following in `fluidPage`)

Layout functions: `tabsetPanel`, `tabPanel`, `navlistPanel`, `navbarMenu`

## Case study

This is one of the first shiny apps I wrote, it pulls data for a specific user
from Stack Overflow, works out which "tags" they have received the most votes
for, then plots those top tags in a word cloud.

### The initial app (app-01.R)

How might this app be improved?

- Code
  - `stackr` eagerly obtains data from stack overflow (maybe it should wait til
  user has input a complete user ID)

- UX
  - Visual style is boring (maybe it should look more like stack overflow?)
  - Could we input user name, rather than ID?
  - Could we show the user-name / image, so the user knows they've copied in the
  correct user ID
  - Can we add some extra plots / tables and make a multipage layout

```{r, code=readLines("examples/06-layout_themes_html-single_to_multipage_app/app-01.R"), eval = FALSE}
```

### The final app (app.R)

Relative to the original app, we have:

- introduced a new results table (containing the data obtained from
  stack-overflow)
- converted from a single-panel to a multi-panel layout (using `tabsetPanel`
  and two `tabPanel`s)
- attempted to use the stack-overflow colour scheme as part of the apps theme,
  and also in the wordcloud colour scheme

The stack overflow colour scheme was obtained by using the browser's developer
tools (here using "Inspect" in chromium and looking at the "styles" section) on
an open SO page.

```{r, code=readLines("examples/06-layout_themes_html-single_to_multipage_app/app-01.R"), eval = FALSE}
```


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/hriFSGeBa_Q")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:05:16	Maya Longhi:	Hello
00:12:32	Russ Hyde:	Hi
00:13:03	Russ Hyde:	I'm just checking my screen share looks OK. Could you tell me if you can read the text on my rstudio instance
00:14:30	Russ Hyde:	Is the text too pale?
00:14:46	Morgan Grovenburg:	No, I can read it
00:14:53	Russ Hyde:	Thanks
00:16:45	Morgan Grovenburg:	I couldn't figure out how to use `bslib::bs_theme_preview()`, will you be showing that today?
00:17:50	Russ Hyde:	I can do if you like
00:19:48	Morgan Grovenburg:	Yes, please! Really, I just need see code that runs it
00:30:02	Layla Bouzoubaa:	@sham, There’s a JavaScript for R book!
00:30:21	shamsuddeen:	Oh, yah Layla
00:30:24	shamsuddeen:	Thanks
00:30:31	Layla Bouzoubaa:	https://book.javascript-for-r.com/
00:30:32	Morgan Grovenburg:	https://book.javascript-for-r.com/
00:38:15	Federica G:	line has gone ?!
00:38:26	Diamantis:	lost the sound but still can see the shared screen
01:05:55	Layla Bouzoubaa:	Sham, Bootstrap is a framework that combines html/js/css that can customized
01:06:06	shamsuddeen:	ok
01:06:11	shamsuddeen:	Thanks Layla
01:07:00	Layla Bouzoubaa:	Like the buttons and stuff in shiny are following bootstrap guidelines (ie the shape, look etc) but you can customize them
01:07:12	shamsuddeen:	ok
01:08:39	Layla Bouzoubaa:	I just discovered this book btw: https://engineering-shiny.org/ . Ch.18.3.3 talks about external css files that may clarify your question on adding “www/“, sham
01:17:51	shamsuddeen:	Oh, Thanks Layla. I got it
01:21:47	Federica G:	Thanks !
01:22:27	shamsuddeen:	Thanks
01:22:38	Morgan Grovenburg:	Thank you Russ!
01:24:22	Anne Hoffrichter:	Thanks Russ!
```
</details>

### Cohort 2

#### Meeting #1 video: `r knitr::include_url("https://www.youtube.com/embed/FN34gu0FzQo")` {-}

<details>
  <summary> Meeting #1 chat log </summary>
  
```
00:07:02	shane:	one sec, dog wants to go see his mom ha
00:07:32	shane:	Kevin, your volume is a little low like it was last week
01:06:06	shane:	team, I unfortunately have to drop. Thank you everyone for the discussion, and thanks to Ryan for presenting and for the great info! If we do end up bleeding into a second session on this, I would definitely be down to be there. Have a great evening!
```
</details>

#### Meeting #2 video: `r knitr::include_url("https://www.youtube.com/embed/W01MoW3AWpY")` {-}

<details>
  <summary> Meeting #2 chat log </summary>
  
```
00:21:58	shane:	https://rstudio.github.io/thematic/
00:25:03	Collin Berke:	https://bootswatch.com/
00:28:50	Conor Tompkins:	"This works because, at plot time, thematic grabs CSS styles from the plot(s) HTML container (via shiny::getCurrentOutputInfo())1 and uses that info to set new R styling defaults."
00:28:56	Conor Tompkins:	From https://rstudio.github.io/thematic/articles/auto.html#shiny
00:31:15	Collin Berke:	Not sure how this relates, but here is the link to the `theme_set()` function I referenced: https://ggplot2.tidyverse.org/reference/theme_get.html
00:46:41	Collin Berke:	You can also see the object here: https://gallery.shinyapps.io/095-plot-interaction-advanced/
00:50:48	Collin Berke:	Here's some more about subsetting list objects. Check out this chapter from Advanced R: https://adv-r.hadley.nz/subsetting.html
01:13:14	Ryan Metcalf:	https://www.osti.gov/servlets/purl/1115145
```
</details>

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/gpj2Uw8imtA")`

<details>
  <summary>Meeting chat log</summary>
```
00:17:24	Njoki Njuki Lucy:	Current deployed slide deck: https://r4ds.github.io/bookclub-mshiny/layout-themes-html.html
00:36:58	Njoki Njuki Lucy:	https://revealjs.com/
01:04:33	Njoki Njuki Lucy:	what is the function cat() doing?
01:10:53	Federica Gazzelloni:	cat {base}: Concatenate and Print
01:11:12	Njoki Njuki Lucy:	Great, thanks! :)
01:11:34	Federica Gazzelloni:	It outputs the objects, concatenating the representations.
01:13:06	Brendan Lam:	I'm also good either way
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/EDcW3IwtBQc")`
