# Uploads and Downloads

In this chapter we learn how to upload/download files to/from a Shiny
 application.
The UI for both uploads and downloads is simple, with essentially one function 
 for each operation.
The server side is more complex.
The upload and download functions for the server each have their own quirks.

## Uploads

### UI

To upload a file use `fileInput(id, label)`. Some other arguments are

- `multiple` - can multiple files be uploaded at once?
- `accept` - what file types are accepted? (character vector)
    - file extension: `.csv`, `.tsv`, `.rds`
    - MIME type: `application/JSON`, `image/png`
    - one of: `audio/*`, `video/*`, `image/*`
    
### Server

In the server, we work with a **data frame** with a special structure.
There are 4 columns:

- `name` - name of the file on user's computer
- `size` - file size in bytes. Default 5 MB; adjust with 
 `shiny.maxRequestSize` option.
- `type` - MIME type of the file
- `datapath` - file path on the server. Temporary.

### Uploading data

There are 2 things to note about uploading a data set.

- Use `req(input$file)` to make sure file is uploaded before code runs.
- Use `accept` argument to `fileInput()` to limit input types.
    - browser doesn't always enforce, so make sure to `validate()`
    
Example of uploading a data set and validating file type (from book):

```{r, eval = FALSE}
ui <- fluidPage(
  fileInput("file", NULL, accept = c(".csv", ".tsv")),
  numericInput("n", "Rows", value = 5, min = 1, step = 1),
  tableOutput("head")
)

server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    
    ext <- tools::file_ext(input$file$name)
    switch(ext,
      csv = vroom::vroom(input$file$datapath, delim = ","),
      tsv = vroom::vroom(input$file$datapath, delim = "\t"),
      validate("Invalid file; Please upload a .csv or .tsv file")
    )
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}
```


## Downloads

### UI

For file downloads, use `downloadButton(id, label)` or `downloadLink(id, label)`.
Customize the appearance with the `class` or `icon` arguments.


### Server

In the server, use `downloadHandler(filename, content)`.
There are only 2 arguments, both are **functions**.

- `filename` - no arguments. Returns file name as a string.
- `content(file)` - one argument (`file`). Path to save the file.

```{r, eval = FALSE}
output$download <- downloadHandler(
  filename = function() {
    paste0(input$dataset, ".csv")
  },
  content = function(file) {
    write.csv(data(), file)
  }
)
```


### Downloading reports

We can allow the user to download a report based on their exploration of an
 application.
 
- parameterized R Markdown is a good way to do this
- potential parameters - filters, simulation parameters
- specify parameters in YAML header of R Markdown document
- **key idea** - call `rmarkdown::render()` from `content` argument of
 `downloadHandler()`

Some tips & tricks:

- `.Rmd` renders in working directory so copy file temporary directory before
 rendering.
- `.Rmd` renders in current R process so consider running in separate session
 with e.g. `callr` package.

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/o157ZBd0CUY")`

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/XXXX")`

<details>
  <summary> Meeting chat log </summary>
  
```
```
</details>
