# Tidy evaluation


**Learning objectives:**

- The difference between a data-variable and an env-variable.
- How to create Shiny apps that lets the user choose which variables will be fed into tidyverse functions like `dplyr::filter()` and `ggplot2::aes()`.

## Tidy Evaluation 

Tidy evaluation is used throughout the tidyverse to make interactive data exploration more fluid, but it comes with a cost: it’s hard to refer to variables indirectly, and hence harder to program with.

In this chapter, you’ll learn how to wrap ggplot2 and dplyr function in a Shiny app. The techniques for wrapping ggplot2 and dplyr functions in a functions and package, are a a little different and covered in other resources like [Using ggplot2 in packages](https://ggplot2.tidyverse.org/dev/articles/ggplot2-in-packages.html) or [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html).

```{r 12-01}
library(shiny)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
```


##  Motivation

Here we create an app that allows you to filter a numeric variable to select rows that are greater than a threshold. The app runs without error, but it doesn’t return the correct result — all the rows have values of carat less than 1. The goal of the chapter is to help you understand why this doesn’t work, and why dplyr thinks you have asked for `filter(diamonds, "carat" > 1)`.

```{r 12-02, eval=FALSE, message=FALSE, warning=FALSE}
num_vars <- c("carat", "depth", "table", "price", "x", "y", "z")
ui <- fluidPage(
  selectInput("var", "Variable", choices = num_vars),
  numericInput("min", "Minimum", value = 1),
  tableOutput("output")
)
server <- function(input, output, session) {
  data <- reactive(diamonds %>% filter(input$var > input$min))
  output$output <- renderTable(head(data()))
}
```

This is a problem of **indirection**: normally when using tidyverse functions you type the name of the variable directly in the function call. But now you want to refer to it indirectly: the variable (`carat`) is stored inside another variable (`input$var`).

- An **env-variable** (environment variable) is a “programming” variables that you create with `<-`. `input$var` is a env-variable.

- A **data-variable** (data frame variables) is “statistical” variable that lives inside a data frame. carat is a data-variable.

With these new terms we can make the problem of indirection more clear: we have a data-variable (carat) stored inside an env-variable (input$var), and we need some way to tell dplyr this. There are two slightly different ways to do this depending on whether the function you’re working with is a “data-masking” function or a “tidy-selection” function.


## Data-masking

Data-masking functions allow you to use variables in the “current” data frame without any extra syntax. It’s used in many dplyr functions like `arrange()`, `filter()`, `group_by()`, `mutate()`, and `summarise()`, and in ggplot2’s `aes()`. Data-masking is useful because it lets you use data-variables without any additional syntax.

### Getting Started

This is a call to `filter()` which uses a data-variable (`carat`) and an env-variable (`min`):

```{r 12-03}
min <- 1
diamonds %>% filter(carat > min)
 
```


This is its base R equivalent:

```{r 12-04}
diamonds[diamonds$carat > min, ]
```

Base R functions refer to data-variables with `$`, and you often have to repeat the name of the data frame multiple times, making it clear what is a data-variable and what is an env-variable. 

It also makes it straightforward to use indirection because you can store the name of the data-variable in an env-variable, and then switch from `$` to `[[`:

```{r 12-05}
var <- "carat"
diamonds[diamonds[[var]] > min, ]
```

We can achieve the same result with tidy evaluation by somehow adding `$` back into the picture while data-masking functions by using `.data` or `.env` to be explicit about whether you’re talking about a data-variable or an env-variable:

```{r 12-06, eval=FALSE}
diamonds %>% filter(.data$carat > .env$min)
```


```{r 12-07, eval=FALSE}
diamonds %>% filter(.data[[var]] > .env$min)
```


```{r 12-08, eval=FALSE, message=FALSE, warning=FALSE}
num_vars <- c("carat", "depth", "table", "price", "x", "y", "z")
ui <- fluidPage(
  selectInput("var", "Variable", choices = num_vars),
  numericInput("min", "Minimum", value = 1),
  tableOutput("output")
)
server <- function(input, output, session) {
  data <- reactive(diamonds %>% filter(.data[[input$var]] > .env$input$min))
  output$output <- renderTable(head(data()))
}
```


Figure 12.2: Our app works now that we’ve been explicit about .data and `.env` and `[[` vs `$`. See live at [https://hadley.shinyapps.io/ms-tidied-up](https://hadley.shinyapps.io/ms-tidied-up).

### Example: ggplot2

Here we apply this idea to a dynamic plot where we allow the user to create a scatterplot by selecting the variables to appear on the `x` and `y` axes. `ggforce::position_auto()` was used so that `geom_point()` works regardless of whether the `x` and `y` variables are continuous or discrete. 

```{r 12-09, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("x", "X variable", choices = names(iris)),
  selectInput("y", "Y variable", choices = names(iris)),
  plotOutput("plot")
)
server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(iris, aes(.data[[input$x]], .data[[input$y]])) +
      geom_point(position = ggforce::position_auto())
  }, res = 96)
}
```


Alternatively, we could allow the user to pick the geom. The following app uses a `switch()` statement to generate a reactive geom that is later added to the plot.

```{r 12-10, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("x", "X variable", choices = names(iris)),
  selectInput("y", "Y variable", choices = names(iris)),
  selectInput("geom", "geom", c("point", "smooth", "jitter")),
  plotOutput("plot")
)
server <- function(input, output, session) {
  plot_geom <- reactive({
    switch(input$geom,
      point = geom_point(),
      smooth = geom_smooth(se = FALSE),
      jitter = geom_jitter()
    )
  })
  
  output$plot <- renderPlot({
    ggplot(iris, aes(.data[[input$x]], .data[[input$y]])) +
      plot_geom()
  }, res = 96)
}
```

Figure 12.3: A simple app that allows you to select which variables are plotted on the `x` and `y` axes. See live at [https://hadley.shinyapps.io/ms-ggplot2](https://hadley.shinyapps.io/ms-ggplot2).

One of the challenges of programming with user selected variables is that your code has to become more complicated to handle all the cases the user might generate.



### Example: dplyr

The same technique also works for dplyr. The following app extends the previous simple example to allow you to choose a variable to filter, a minimum value to select, and a variable to sort by.

```{r 12-11, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("var", "Select variable", choices = names(mtcars)),
  sliderInput("min", "Minimum value", 0, min = 0, max = 100),
  selectInput("sort", "Sort by", choices = names(mtcars)),
  tableOutput("data")
)
server <- function(input, output, session) {
  observeEvent(input$var, {
    rng <- range(mtcars[[input$var]])
    updateSliderInput(
      session, "min", 
      value = rng[[1]], 
      min = rng[[1]], 
      max = rng[[2]]
    )
  })
  
  output$data <- renderTable({
    mtcars %>% 
      filter(.data[[input$var]] > input$min) %>% 
      arrange(.data[[input$sort]])
  })
}
```

Figure 12.4: A simple app that allows you to pick a variable to threshold, and choose how to sort the results. See live at [https://hadley.shinyapps.io/ms-dplyr](https://hadley.shinyapps.io/ms-dplyr).

Most other problems can be solved by combining `.data` with your existing programming skills. For example, what if you wanted to conditionally sort in either ascending or descending order?


```{r 12-12, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("var", "Sort by", choices = names(mtcars)),
  checkboxInput("desc", "Descending order?"),
  tableOutput("data")
)
server <- function(input, output, session) {
  sorted <- reactive({
    if (input$desc) {
      arrange(mtcars, desc(.data[[input$var]]))
    } else {
      arrange(mtcars, .data[[input$var]])
    }
  })
  output$data <- renderTable(sorted())
}
```


### User supplied data

This app allows the user to upload a tsv file, then select a variable and filter by it. It will work for the vast majority of inputs that you might try it with. 

```{r 12-13, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  fileInput("data", "dataset", accept = ".tsv"),
  selectInput("var", "var", character()),
  numericInput("min", "min", 1, min = 0, step = 1),
  tableOutput("output")
)
server <- function(input, output, session) {
  data <- reactive({
    req(input$data)
    vroom::vroom(input$data$datapath)
  })
  observeEvent(data(), {
    updateSelectInput(session, "var", choices = names(data()))
  })
  observeEvent(input$var, {
    val <- data()[[input$var]]
    updateNumericInput(session, "min", value = min(val))
  })
  
  output$output <- renderTable({
    req(input$var)
    
    data() %>% 
      filter(.data[[input$var]] > input$min) %>% 
      arrange(.data[[input$var]]) %>% 
      head(10)
  })
}
```


Figure 12.5: An app that filter users supplied data, with a surprising failure mode See live at [https://hadley.shinyapps.io/ms-user-supplied](https://hadley.shinyapps.io/ms-user-supplied).

It'll work with the vast majority of data frames. However, if the data frame contains a variable called `input`, we get an error message because `filter()` is attempting to evaluate `df$input$min`:

```{r 12-14, eval=FALSE}
df <- data.frame(x = 1, y = 2)
input <- list(var = "x", min = 0)

df %>% filter(.data[[input$var]] > input$min)

```


```{r 12-15, eval=FALSE}
df <- data.frame(x = 1, y = 2, input = 3)
df %>% filter(.data[[input$var]] > input$min)

```

This problem is due to the ambiguity of data-variables and env-variables, and because data-masking prefers to use a data-variable if both are available. We can resolve the problem by using `.env` to tell `filter()` only look for min in the env-variables:

```{r 12-16, eval=FALSE}
df %>% filter(.data[[input$var]] > .env$input$min)
```


You only need to worry about this problem when working with user supplied data; when working with your own data, you can ensure the names of your data-variables don’t clash with the names of your env-variables.

### Why not use base R?

You might wonder if you’re better off without filter(), and if instead you should use the equivalent base R code:

```{r 12-17, eval=FALSE}
df[df[[input$var]] > input$min, ]
```


That’s a totally legitimate position, as long as you’re aware of the work that filter() does for you so you can generate the equivalent base R code. In this case:

- You’ll need `drop = FALSE` if df only contains a single column (otherwise you’ll get a vector instead of a data frame).

- You’ll need to use `which()` or similar to drop any missing values.

- You can’t do group-wise filtering (e.g. `df %>% group_by(g) %>% filter(n() == 1)`).

In general, if you’re using dplyr for very simple cases, you might find it easier to use base R functions that don’t use data-masking. However, in my opinion, one of the advantages of the tidyverse is the careful thought that has been applied to edge cases so that functions work more consistently.

## Tidy-selection

As well as data-masking, there’s one other important part of tidy evaluation: tidy-selection. Tidy-selection provides a concise way of selecting columns by position, name, or type. It’s used in `dplyr::select()` and `dplyr::across()`, and in many functions from tidyr, like `pivot_longer()`, `pivot_wider()`, `separate()`, `extract()`, and `unite()`.


### Indirection

To refer to variables indirectly use `any_of()` or `all_of()`: both expect a character vector env-variable containing the names of data-variables. The only difference is what happens if you supply a variable name that doesn’t exist in the input: `all_of()` will throw an error, while `any_of()` will silently ignore it.


```{r 12-18, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("vars", "Variables", names(mtcars), multiple = TRUE),
  tableOutput("data")
)

server <- function(input, output, session) {
  output$data <- renderTable({
    req(input$vars)
    mtcars %>% select(all_of(input$vars))
  })
}
```


### Tidy Selection and Data Masking


Working with multiple variables is trivial when you’re working with a function that uses tidy-selection: you can just pass a character vector of variable names into `any_of()` or `all_of()`. Wouldn’t it be nice if we could do that in data-masking functions too? That’s the idea of the `across()` function, added in dplyr 1.0.0. It allows you to use tidy-selection inside data-masking functions.
`across()` is typically used with either one or two arguments. The first argument selects variables, and is useful in functions like `group_by()` or `distinct()`. 


```{r 12-19, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("vars", "Variables", names(mtcars), multiple = TRUE),
  tableOutput("count")
)

server <- function(input, output, session) {
  output$count <- renderTable({
    req(input$vars)
    
    mtcars %>% 
      group_by(across(all_of(input$vars))) %>% 
      summarise(n = n(), .groups = "drop")
  })
}
```


Figure 12.6: This app allows you to select any number of variables and count their unique combinations. See live at [https://hadley.shinyapps.io/ms-across](https://hadley.shinyapps.io/ms-across).

The second argument is a function (or list of functions) that’s applied to each selected column. That makes it a good fit for `mutate()` and `summarise()` where you typically want to transform each variable in some way. For example, the following code lets the user select any number of grouping variables, and any number of variables to summarise with their means.

```{r 12-20, eval=FALSE, message=FALSE, warning=FALSE}
ui <- fluidPage(
  selectInput("vars_g", "Group by", names(mtcars), multiple = TRUE),
  selectInput("vars_s", "Summarise", names(mtcars), multiple = TRUE),
  tableOutput("data")
)

server <- function(input, output, session) {
  output$data <- renderTable({
    mtcars %>% 
      group_by(across(all_of(input$vars_g))) %>% 
      summarise(across(all_of(input$vars_s), mean), n = n())
  })
}
```

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/Z3byy6Z73P4")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:01:39	Russ Hyde:	Hi, am I audible?
00:09:57	Russ Hyde:	Is the connection OK for everyone?
00:09:59	priyanka gagneja:	I am not hearing you so well right now Robert
00:10:03	Andrew MacDonald:	I might be losing sound?
00:10:08	Diamantis:	I have sound issues, can't hear clearly
00:10:14	Andrew MacDonald:	ah seems to be just on Rob?
00:10:35	andrew bates:	I didn’t have sound. Had to call in.
00:11:00	Andrew MacDonald:	In the meantime I’d just like to say hello!
00:11:09	Andrew MacDonald:	This is my first time attending!
00:11:13	Russ Hyde:	Hello Andrews
00:11:16	andrew bates:	Welcome!
00:11:23	Andrew MacDonald:	we have an abundance of Andrews :)
00:11:43	andrew bates:	yeah!
00:12:01	Andrew MacDonald:	I’m an ecologist who develops some shiny apps on a contract — a complete convert to the Way of Golem & v excited for Colin’s appearance next week
00:12:48	Russ Hyde:	Cool. I'm really pleased we were able to get him to talk
00:13:07	priyanka gagneja:	Nice to have you Andrew
00:14:00	Russ Hyde:	Have any of you been attending the other bookclubs?
00:14:17	Andrew MacDonald:	(there are other book clubs?)
00:14:18	priyanka gagneja:	I do if my schedule allows
00:14:33	Anne Hoffrichter:	I was in Advanced R bc
00:14:38	priyanka gagneja:	Yes there are… and they are now increasing at crazy pace
00:14:41	andrew bates:	Shameless self advocacy: if they’re looking for more Shiny devs, I’m looking :)
00:15:00	Russ Hyde:	(ooo that was pretty neat)
00:15:46	Anne Hoffrichter:	nice, i didn’t know you can run a shiny app from an rmd chunk!
00:16:23	Andrew MacDonald:	that is a cool way to show this! I thought maybe that was the default from the group
00:17:00	Andrew MacDonald:	could I ask what is the typical format here? do we begin with a summary and then discuss, or do we ask questions during Rob’s description?
00:17:42	Russ Hyde:	Feel free to ask questions whenever.
00:19:55	Andrew MacDonald:	"foo" > 42
00:20:41	priyanka gagneja:	@anne - Yeah you can add shiny code inside a Rmd. When you add shiny code and save it , it shows a ‘Run Document’ option instead of Run App.
00:28:42	Andrew MacDonald:	okay I’ll be honest i’m just scared of <<-
00:29:06	priyanka gagneja:	Less than
00:29:13	Andrew MacDonald:	+ 1 :)
00:31:51	Russ Hyde:	I try to avoid <<- as well.
00:32:08	docksbox@pm.me:	Yes
00:32:14	priyanka gagneja:	Good point. So you mean as a crux of it , it would help the most for debugging ?
00:32:54	priyanka gagneja:	If so, I am ready to start using it then
00:33:11	Russ Hyde:	f <- function(){counts= 0; function(){counts <<- counts + 1; message(counts)}}
00:33:22	Russ Hyde:	g <- f(); g(); g()
00:33:45	Russ Hyde:	That should print 1 then 2
00:35:49	Russ Hyde:	This is a really neat way to debug
00:36:17	Russ Hyde:	(I meant Robert's example; not the hack I just posted)
00:37:01	Andrew MacDonald:	laptop battery died! I’m back :)
00:41:28	Andrew MacDonald:	tbh I strongly dislike the user-supplied data example! the word “data” is in there but i think it means.. three things? the input, the reactive expression, and the .data thing
00:42:18	Russ Hyde:	Haha, I just wrote a chapter about good-names in R programming. Calling your data 'data' is really not a good idea ...
00:42:42	Andrew MacDonald:	exactly!!!
00:47:10	priyanka gagneja:	Reminds me of days when I did parse eval :’(
00:51:13	Robert Overman:	https://adv-r.hadley.nz/quasiquotation.html
00:52:15	Robert Overman:	https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html
00:57:39	Andrew MacDonald:	very slick!!!
00:57:44	Andrew MacDonald:	(get it)
00:59:38	Russ Hyde:	select(df, A, B, C) vs d <- c("A", "B", "C"); select(df, !!!d)
00:59:41	Russ Hyde:	?
01:00:28	priyanka gagneja:	aah
01:00:37	priyanka gagneja:	Yeah .. I think so .. named df
01:00:43	priyanka gagneja:	Named vector sorry
01:01:04	Anne Hoffrichter:	I gotta run, thanks Robert. See you all next week.
01:01:09	Robert Overman:	final_colnames <- data.frame(
  stringsAsFactors = FALSE,
  new_nm = c(
    "A","B","C"
  ),
  old_nm = c(
    "X","Y","Z"
  )
)
new_nm <- final_colnames$new_nm %>% unlist()
old_nm <- final_colnames$old_nm %>% unlist()
names(old_nm) <- new_nm

old_data <- data.frame(X = 1, Y = 1, Z = 1)
old_data %>% select(!!!old_nm)
01:01:25	priyanka gagneja:	gotcha
01:01:35	Andrew MacDonald:	I must go as well — thank you everyone! nice to join this effort!
01:01:48	priyanka gagneja:	I recall I did this recently .. n I wasn’t sure why one worked n not other one.
01:05:34	shamsuddeen:	Thanks Robert
01:05:43	Andrew MacDonald:	yeah thanks Rob!
```
</details>

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/EGu10YO_Bn4")`

<details>
  <summary> Meeting chat log </summary>
  
```
01:11:18	Ryan Metcalf:	https://databank.worldbank.org/reports.aspx?source=global-economic-prospects
```
</details>


### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/wrN7Xv85dWM")`

<details>
  <summary>Meeting chat log</summary>
```
00:55:54	Federica Gazzelloni:	VROOM: https://www.rdocumentation.org/packages/vroom/versions/1.0.2/topics/vroom
00:56:22	Federica Gazzelloni:	in practice: vroom: Read a delimited file into a tibble
```
</details>


<!--
### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary>Meeting chat log</summary>
```
LOG
```
</details>
-->
