# (PART\*) Mastering reactivity {.unnumbered}

# Why reactivity?

This chapter starts a new section of the book focused on reactivity.
Yes, we have already learned about reactivity in earlier chapters.
But in this section, we shift from an intuitive understanding based on applying
 reactivity to a formal understanding of the underlying theory.
This chapter focuses on what reactivity, or reactive programming, is and why
 it is used in Shiny.
 
## Introduction

- Shiny is "magic"
- specifically, it is "good" magic
- Shiny magic = "simple concepts combined in consistent ways" (vs. heuristics
 and special cases)
 
## Why do we need reactive programming?
 
What is reactive programming?
 
 - "focuses on values that change over time, and calculations and actions that
  depend on those values"
- ([Wikipedia](https://en.wikipedia.org/wiki/Reactive_programming)) "... 
 reactive programming is ... concerned with data streams and the 
 **propagation of changes** (emphasis mine).
 
Why reactive programming in Shiny?

- Shiny applications are interactive
- need something dynamic, unlike most R code
- want inputs & outputs to stay in sync, while minimizing computation
  - outputs and reactive expressions change _if and only if_ their inputs change
- reactive programming -> **automatic** updating/propagation of **necessary**
 changes

### Why can't you use variables?

- reactive programming: values that change over time.
- don't variables change over time?
- variables don't update automatically -> no good

```{r why-not-variables, eval = FALSE}
temp_c <- 10
temp_f <- (temp_c * 9 / 5) + 32
temp_f
#> [1] 50

temp_c <- 30

# temp_f not automatically updated
temp_f
#> [1] 50
```



### Why about functions?

- solves automatic updating problem
- doesn't solve necessary computation problem

```{r what-about-functions, eval = FALSE}
temp_c <- 10
temp_f <- function() {
  message("Converting") 
  (temp_c * 9 / 5) + 32
}
temp_f()
#> Converting
#> [1] 50

# temp_f automatically updated
temp_c <- -3
temp_f() 
#> Converting
#> [1] 26.6

# but does unnecessary computation (recomputes every time it's called)
temp_f() 
#> Converting
#> [1] 26.6
```


### Event-driven programming

- before, Shiny would probably have used event-driven programming
  - callback functions that run in response to events
  - Ex: when button clicked, run function to process order
- can implement with R6 classes in R

```{r event-driven-programming-class-def, eval = FALSE}
DynamicValue <- R6::R6Class("DynamicValue", list(
  value = NULL,
  on_update = NULL,
  
  get = function() self$value,

  set = function(value) {
    self$value <- value
    if (!is.null(self$on_update)) 
      self$on_update(value)
    invisible(self)
  },
  
  onUpdate = function(on_update) {
    self$on_update <- on_update
    invisible(self)
  }
))
```


```{r event-driven-programming-class-usage, eval = FALSE}
temp_c <- DynamicValue$new()
temp_c$onUpdate(function(value) {
  message("Converting") 
  temp_f <<- (value * 9 / 5) + 32
})

temp_c$set(10)
#> Converting
temp_f
#> [1] 50

temp_c$set(-3)
#> Converting
temp_f
#> [1] 26.6
```


- event-driven programming solves problem of unnecessary computation
- but introduces a new problem
  - very carefully track which inputs affect which computations
  - hard to balance correctness vs. performance
  
## Reactive programming

Reactive programming combines features of solutions we've seen so far.

```{r reactive-programming, eval = FALSE}
library(shiny)
reactiveConsole(TRUE) # let's us use reactivity in console

temp_c <- reactiveVal(10) # create
temp_c()                  # get
#> [1] 10
temp_c(20)                # set
temp_c()                  # get
#> [1] 20


temp_f <- reactive({
  message("Converting") 
  (temp_c() * 9 / 5) + 32
})
temp_f()
#> Converting
#> [1] 68

# temp_f automatically updates
temp_c(-10)
temp_f()
#> Converting
#> [1] 14

# _and_ only computes when needed (don't see "Converting")
temp_f()
#> [1] 14
```


2 important properties of reactive expressions

- **lazy**. only does work when it's called
- **cached**. saves last result -> only does work on first call

## A brief history of reactive programming

- spreadsheets! (VisCalc, 1979)
- not really studied academically until late 1990s
- 2010s -> mainstream
  - JavaScript UI frameworks: Knockout, Ember, Meteor
  - now: React, Vue, Angular
- "reactive programming" is a general term. a lot of variablility in 
 implementations and terminology


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/tXShS1PVKlY")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:01:19	Russ Hyde:	Hi Everyone4
00:05:22	Russ Hyde:	There are lots of chapters available if anyone would like to do a talk over the next few weeks.
00:05:37	Russ Hyde:	Myself and Diamantis have the next couple of weeks planned.
00:06:16	Russ Hyde:	But there's another chapter in the Reactivity section and all-but-one of the best-practices chapters available
00:06:50	Andrew MacDonald (he/him):	hello everyone! :
00:06:53	Andrew MacDonald (he/him):	:D
00:06:57	Russ Hyde:	Those chapters take us up to mid-august.
00:07:13	Russ Hyde:	And hopefully we'll all be expert shiny devs by then ...
00:09:14	Layla Bouzoubaa:	I think I can do functions
00:09:19	Layla Bouzoubaa:	*think*
00:10:44	Andrew MacDonald (he/him):	I could be an understudy for ch16 :)
00:12:03	Russ Hyde:	Layla, if you'd like to do functions that would be great
00:12:33	Layla Bouzoubaa:	Ok, I’ll go ahead and put it in the books
00:12:39	Layla Bouzoubaa:	6/29
00:12:55	Russ Hyde:	:)
00:31:55	priyanka gagneja:	Yeah {react log}
00:41:21	priyanka gagneja:	@Layla - you might like this https://www.youtube.com/watch?v=8VjkatOf7Bs
00:48:49	Layla Bouzoubaa:	Sorry everyone, I need to hop off. See you next week!!
00:49:31	priyanka gagneja:	By Maya Gans :P
00:50:09	priyanka gagneja:	Well .. I meant author of the book ‘javascript for R’ that AB brought up
00:52:08	Andrew MacDonald (he/him):	this was a lighter chapter! touches on a lot of topics though
00:52:20	Andrew MacDonald (he/him):	yeah thanks Andrew B., great summary
```
</details>

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/XXXX")`

<details>
  <summary> Meeting chat log </summary>
  
```
```
</details>
