# Reactive Building Blocks

## Learning Objectives

- Building blocks:
  - Reactive values
  - Reactive expressions
  - Observers

- How these tools are built from low-level functions:
  - `observe()`
  - `isolate()`
  
- How error messages / signal conditions move on reactive graph

- Shiny reactive values are built on reference-semantics

## Tools

- Timed invalidation
- `reactiveVal(some_value)` and `reactiveValues(a = abc, b = xyz)`
- `reactiveConsole(TRUE)`
- `{reactlog}`

## Prerequisite Knowledge

- Important properties of reactives:
  - lazy
  - cached

- (non-shiny) R things
  - Reference- vs Copy-On-Modify semantics
  - Error handling in R
  - Functions that are used inside other functions (on.exit, match.args)

- Shiny examples and functions:
  - difference between usage of (Section 3.5 and 3.6)
    - `eventReactive` --- you want the value
    - and `observeEvent` --- you want a side-effect
  - `shiny::req()` (Chapter 8)
  - `shiny::reactiveTimer()` (Section 3.5.1)

## Reactive Values

```{r, echo = FALSE}
rv_details <- tibble::tribble(
  ~ "", ~ "reactiveVal()", ~ "reactiveValues()",
  "Holds", "Single value", "Multiple values (sim to list)",
  "Definition", "x <- reactiveVal(1)", "x <- reactiveValues(a = 1, b = 2)",
  "Getting syntax", "x()", "x\\$a",
  "Setting syntax", "x(2)", "x\\$a <- 2",
  "Class", "reactiveVal,reactive,function", "reactivevalues"
)
```

```{r, echo = FALSE}
rv_details %>%
  kableExtra::kbl() %>%
  kableExtra::kable_styling()
```

Reference semantics

```{r}
# Most of R uses copy-on-modify semantics
f <- function(x) {
  x$a = 2
  x
}
x <- list(a = 1, b = 2)
```

```{r}
f(x) # a new object is returned
```

```{r}
x # ... and the initial object is unchanged
```

```{r}
x1 <- x2 <- 1
x2 <- 2
x1
x2
```

```{r}
# R6 classes use reference semantics
g <- function(x) {
  x$a = 2
  invisible(NULL)
}
Y <- R6::R6Class("Y", list(a = 1, b = 2))
y <- Y$new()
y
```

```{r}
g(y)
y
```

```{r}
y1 <- y2 <- Y$new()
y2$a <- 2
y1
y2
```

Both `reactiveVal` and `reactiveValues` have reference semantics

```r
# Exercise from the book
# Run this interactively:

l1 <- reactiveValues(a = 1, b = 2)
l2 <- list(a = reactiveVal(1), b = reactiveVal(2))

# ... with a little extra
l3 <- reactiveVal(list(a = 1, b = 2))

# Getting
l1$a
l1[["a"]]
l2$a()
l2[["a"]]()
l3()$a

# Setting
l1$a <- 15
l2$a(15)
# For l3, can't easily update just 'a'
l3(list(a = 15, b = 2))
```

```r
# Exercise, run this interactively:

# reactiveVal has reference semantics
x <- reactiveVal(1)
y <- x
z <- reactiveVal(1)

x_init_address <- pryr::address(x)

x()
y()
z()

x(2)

x()
y()
z()

pryr::address(x) == pryr::address(y)
pryr::address(x) == pryr::address(z)
pryr::address(x) == x_init_address
```

## Reactive Expressions

Example of a reactive expression from Section 8.2.2 (User Feedback, Notifications, Removing on
completion)

```r
server <- function(input, output, session) {
  data <- reactive({
    # Several lines of code can go inside this expression
    id <- showNotification("Reading data...", duration = NULL, closeButton = FALSE)
    
    # the code inside `on.exit(...)` runs after this whole expression is evaluated
    on.exit(removeNotification(id), add = TRUE)
    
    # on success, this value is cached in `data()`
    read.csv(input$file$datapath)
  })
  
  # ... some code that uses data()
}
```

### Errors

Errors
- are cached just like values
- propagate through the reactive graph
- different behaviour when they reach
  - output (error is displayed by the app)
  - observer (eg, `observeEvent()`; crashes the session; unless you use try/tryCatch)

```r
# Errors are cached

# Run this in the console
r <- reactive(stop("Error occured at ", Sys.time(), call. = FALSE))
r()
Sys.sleep(2)
r()
```

```r
# Errors propagate through the reactive graph

# Exercise --- run this in the console
ui <- fluidPage(
  checkboxInput("error", "error?"),
  textOutput("result")
)
server <- function(input, output, session) {
  a <- reactive({
    if (input$error) {
      stop("Error!")
    } else {
      1
    }
  })
  b <- reactive(a() + 1)
  c <- reactive(b() + 1)
  output$result <- renderText(c())
}

reactlog::reactlog_enable()
shinyApp(ui, server)
```

### `on.exit()`

Where else do you see `on.exit()`?

- inside [function bodies](https://adv-r.hadley.nz/functions.html#on-exit) (eg, close this file once
the function completes, reset the plotting options after making this graph)
- in [testthat expressions](https://www.tidyverse.org/blog/2020/04/self-cleaning-test-fixtures/)

In functions, for example, the code in `on.exit()` runs after all the rest of the code has run.
It runs even if there are errors / warnings.
You can have multiple calls to `on.exit()` inside a function (use add = TRUE, so a call doesn't overwrite an earlier one)

The expression in the definition of a `reactive` can be thought of as a function body (but with
automatic caching / laziness):

```r
# Think of this:
reactive({
  do_something(x(), y())
})
# or eventReactive(triggers, {do_something(x(), y())})

# Like this:
function() {
  do_something(x(), y())
}
```

Because that's basically what it is:

```{r}
# note that `fun <- as_function(x)` line ...
reactive
```


## Observers and Outputs

Reactives : cached and lazy

Observers / Outputs : forgetful and eager

Why does `x()` never run here?

```r
server <- function(input, output, session) {
  x <- reactive({
    rnorm(10)
  })
}
```

In the reactive graph, observers and outputs are the terminal nodes.
If they were lazy, nothing would get done.

Eagerness is infectious

Observer:

- side-effect node (eg, write to file, send messages to console)
- usually defined using `observeEvent()`

Output:

- as assigned to `output$x <- ...` variable

Both created using low-level function `observe()`

```r
# observers are eager

y <- reactiveVal(10)
observe({
  message("`y` is ", y())
})
y(5)
y(4)
```

`observe()` doesn't do something, it creates something

```r
# example from the book (I found this a bit confusing)
x <- reactiveVal(1)
y <- observe({
  x()
  observe(print(x()))
})
x(2)
x(3)

# y() reacts each time x() updates
# each time y() reacts it adds a new observer node that reacts to x()
```

```r
# alternative example:
# each time you change f, a new observer of g is added to the graph
f <- reactiveVal(1)
g <- reactiveVal(2)
h <- observe({
  f()
  observe(print(g()))
})
```

## Isolating Code

`observeEvent()` and `eventReactive` allow you to access values in a reactive without taking a
reactive dependency on them.

How would you write a function to achieve that?

From section 3.5.2

```r
ui <- fluidPage(
  fluidRow(
    column(3, 
      numericInput("lambda1", label = "lambda1", value = 3),
      numericInput("lambda2", label = "lambda2", value = 5),
      numericInput("n", label = "n", value = 1e4, min = 0),
      actionButton("simulate", "Simulate!")
    ),
    column(9, plotOutput("hist"))
  )
)

server <- function(input, output, session) {
  # x1 depends on `simulate`, `n` and `lambda1`
  # ... but only has a "reactive dependency" on `simulate`
  # ... sorry, ... what?
  
  x1 <- eventReactive(input$simulate, {
    rpois(input$n, input$lambda1)
  })
  x2 <- eventReactive(input$simulate, {
    rpois(input$n, input$lambda2)
  })

  output$hist <- renderPlot({
    freqpoly(x1(), x2(), binwidth = 1, xlim = c(0, 40))
  }, res = 96)
}
```

So we've got

```r
new_reactive <- eventReactive(dependOnThis, runThis)
```

`isolate(expression)` is used.

- separates access from dependency
- executes the expression in a scope where the reactive values can be read (see `? isolate`)
- reduces the time the reactive graph is invalidated

Counting example from the book:

```r
r <- reactiveValues(count = 0, x = 1)
observe({
  r$x
  # without the isolate() call, an infinite loop would occur
  r$count <- isolate(r$count) + 1
})

r$x <- 1
r$x <- 2
r$count
```

In `eventReactive(dependOnThis, runThis)`,
if you follow the code:
- eventReactive calls `bindEvent()`
- bindEvent (basically) calls `reactive({dependOnThis; isolate(runThis)})`

So 
- eventReactive(x, y) ===~ reactive(x, isolate(y))
- (and similarly) observeEvent(x, y) ===~ observe({x, isolate(y)})

## Timed Invalidation

In section 3.5.1

```r
server <- function(input, output, session) {
  # timer is invalidated every 500ms
  # A) How is that possible?
  timer <- reactiveTimer(500)
  
  x1 <- reactive({
    # ... hence, a new x1 and x2 are sampled each 500ms
    # B) What if sampling x1 took longer than 500ms?

    timer()
    rpois(input$n, input$lambda1)
  })
  x2 <- reactive({
    timer()
    rpois(input$n, input$lambda2)
  })
  
  output$hist <- renderPlot({
    freqpoly(x1(), x2(), binwidth = 1, xlim = c(0, 40))
  }, res = 96)
}
```

`invalidateLater(ms)` allows the programmer to invalidate a node in the graph

```r
# Example, add a new random number to a sum once every 300ms

sum <- reactiveVal(0)
observe({
  invalidateLater(300)
  sum(isolate(sum()) + runif(1))
})
```

Note that `invalidateLater` runs from the time it is invoked.
Combine it with on.exit() to ensure the node is invalidated a fixed-time after the node's
computation completes

```r
# e.g.: iterate: load a file, one second after you finished reading it

data <- reactive({
  on.exit(invalidateLater(1000))
  read.csv("data.csv")
})
```

Additional tools `reactivePoll` and `reactiveFileReader` are built on invalidateLater

Beware, the accuracy of timing may be affected by other processes


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/uwawZNrU-5k")`

<details>
  <summary> Meeting chat log </summary>
  
```
01:04:02	Andrew MacDonald:	I’m afraid i ahve to duck out early today. Please let me know if we still need somebody for next week!
01:04:09	Andrew MacDonald:	thanks so much Russ et al! :D
01:04:18	docksbox@pm.me:	thanks
01:04:40	docksbox@pm.me:	are there any example apps ?
01:11:12	Layla Bouzoubaa:	Thanks russ! Need to hop off!
01:14:19	Anne Hoffrichter:	Thanks Russ! See you next week!
01:14:34	russ:	Bye
```
</details>

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/q2d3uBHO3Tk")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:37:01	Ryan Metcalf:	Really good topic on Linked List in C++. The discussion was related to memory management. https://www.geeksforgeeks.org/data-structures/linked-list/
00:45:12	Kevin Gilds:	Reactive Values in Modules
00:45:39	Kevin Gilds:	The inputs across across modules
00:47:27	Kevin Gilds:	I will put some articles in the slack reactive values and modules
```
</details>


<!--
### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary>Meeting chat log</summary>
```
LOG
```
</details>
-->


<!--
### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary>Meeting chat log</summary>
```
LOG
```
</details>
-->
