# (Best practices) General guidelines

Big apps or big teams == New challenges

- Organisation

- Stability

- Maintainability

## "Best practices" section of the book

- Software engineering approach & mindset (today's chapter)

- Decomposition:

  - into functions

  - into modules
  
- Organisation

  - into packages
  
- Stability

  - Testing

- Security

- Performance

## Software engineering

```{r, echo = FALSE}
library(DiagrammeR)
```

```{r, echo = FALSE}
mermaid("
graph LR
A[\"I don't understand it\"]-->B[\"I still need the docs\"]
B-->C[\"I understand it fluently\"]
")
```

## Code organization

Empathy

Code clarity

Comments

Function size

Copy/Paste development

Shiny: functions, modules

## Testing

What is a "test plan"

Why does testing improve stability of an app

Automation vs Manual

Test types:

- Unit, Integration, Functional, Load ....

- Testing behaviour vs Testing components

Can you have _too many_ tests

Can you have the _wrong kind_ of tests

## Dependency management

Reproducibility

Environment management: `{renv}`, `{packrat}`, conda, docker, `sessionInfo()`

Configuration: `{config}`

Are there different reproducibility requirements for

- an app vs
- an analysis project?

When should you have different configurations for an app?

The filesystem as part of the environment

## Source code management

R uses svn: [https://github.com/r-devel/r-svn#readme]()

Did anyone use 'git' for the first time during the bookclub

Related tools for version control-type things (tracking changes etc)

Anyone using version-control systems other than git?

What are the benefits and pain-points of git / github etc?

## Continuous integration / Continuous deployment

Continuous integration for an app

- what things would you run?

The bookclub as an example of continuous deployment

## Code reviews

A fresh pair of eyes

Speeds up learning (try contributing to an R package, and see how much you don't yet know)

Tricks to speed up code-review?

Who should review who?

Code review antipatterns?

Alternatives / complements? pair programming