# General guidelines

Big apps or big teams == New challenges

- Organisation

- Stability

- Maintainability

## "Best practices" section of the book

- Software engineering approach & mindset (today's chapter)

- Decomposition:

  - into functions

  - into modules
  
- Organisation

  - into packages
  
- Stability

  - Testing

- Security

- Performance

## Software engineering

```{r, echo = FALSE}
library(DiagrammeR)
```

```{r, echo = FALSE}
mermaid("
graph LR
A[\"I don't understand it\"]-->B[\"I still need the docs\"]
B-->C[\"I understand it fluently\"]
")
```

## Code organization

Empathy

Code clarity

Comments

Function size

Copy/Paste development

Shiny: functions, modules

## Testing

What is a "test plan"

Why does testing improve stability of an app

Automation vs Manual

Test types:

- Unit, Integration, Functional, Load ....

- Testing behaviour vs Testing components

Can you have _too many_ tests

Can you have the _wrong kind_ of tests

## Dependency management

Reproducibility

Environment management: `{renv}`, `{packrat}`, conda, docker, `sessionInfo()`

Configuration: `{config}`

Are there different reproducibility requirements for

- an app vs
- an analysis project?

When should you have different configurations for an app?

The filesystem as part of the environment

## Source code management

R uses svn: [https://github.com/r-devel/r-svn#readme]()

Did anyone use 'git' for the first time during the bookclub

Related tools for version control-type things (tracking changes etc)

Anyone using version-control systems other than git?

What are the benefits and pain-points of git / github etc?

## Continuous integration / Continuous deployment

Continuous integration for an app

- what things would you run?

The bookclub as an example of continuous deployment

## Code reviews

A fresh pair of eyes

Speeds up learning (try contributing to an R package, and see how much you don't yet know)

Tricks to speed up code-review?

Who should review who?

Code review antipatterns?

Alternatives / complements? pair programming


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/stfuGTNJD2M")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:06:04	russ:	Hi everyone
00:06:14	russ:	Hopefully we can have a discussion today
00:06:35	russ:	I've put some notes into the bookclub book, but haven'thad them merged yet
00:12:38	andrew bates:	https://blog.rstudio.com/2021/03/11/time-to-shiny/
00:53:28	russ:	cache?
00:53:52	andrew bates:	https://shiny.rstudio.com/app-stories/weather-lookup-caching.html
01:04:33	Federica:	thanks
```
</details>

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/aQHyWOEvJAI")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:27:24	Kevin Gilds:	https://www.rostrum.blog/2021/04/14/gha-readme/
00:27:42	Kevin Gilds:	Update README with Github actions
00:51:26	collinberke:	https://www.youtube.com/watch?v=BEyiFVWaTtc&list=PL3x6DOfs2NGjtn1_4BSX99R5wrLjK7XvY&index=19
00:51:35	collinberke:	https://www.youtube.com/watch?v=tWEa6JwlJ3c&list=PL3x6DOfs2NGjtn1_4BSX99R5wrLjK7XvY&index=21
```
</details>


### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/JdisqyPzfHo")`



### Cohort 4

`r knitr::include_url("https://www.youtube.com/embed/7b5UQvzgPxo")`

<details>
  <summary>Meeting chat log</summary>
```
00:22:45	Trevin Flickinger:	Hi everyone 👋
00:22:58	Lydia Gibson:	Hello
00:28:45	Trevin Flickinger:	No questions
00:28:48	Lucio Cornejo:	all good so far
00:32:54	Trevin Flickinger:	yes
00:36:03	Trevin Flickinger:	I can see it now
00:36:11	Trevin Flickinger:	Sorry I can only type today
00:44:31	Matthew Efoli:	https://dcl-prog.stanford.edu/purrr-parallel.html#:~:text=The%20pmap()%20functions%20work,to%20supply%20to%20your%20function.
00:59:07	Lucio Cornejo:	session = getReactiveDomain()
01:00:04	Lucio Cornejo:	session = getDefaultReactiveDomain()
01:01:04	Trevin Flickinger:	We’ll get more into sessions next week with modules
01:01:55	Lucio Cornejo:	same over here
01:03:16	Trevin Flickinger:	Thanks Matthew! Nothing from me
01:03:41	Trevin Flickinger:	See everyone next week 😄
01:03:41	Lucio Cornejo:	Also excited for next week. Thank you, see you next everyone
```
</details>


### Cohort 5

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary>Meeting chat log</summary>
```
LOG
```
</details>
