# Functions


Like in R analyses, functions can be useful in Shiny applications to improve the functionality/optimization and make your code cleaner for quicker debugging. 

**Learning objectives:**

+ Learn how to use functional programming techniques to make many controls at once

+ Understand how to use functions with reactives in the server

+ Walk through some use cases where functions may be appropriate

------------------------------------------------------

Outline

1 - File Organization

2 - UI functions

3 - Server functions

----------------------------------------------------------

## File Organization 

+ *Functions* allow you to spread your code across multiple files
  + this is where organization is `r emo::ji("key")`
+ Make sure to have an `R/` directory and in it, place large functions into their own script - `R/myFunc.R`
  + Collections of smaller functions can go into 1 script, i.e. `R/utils.R`
  + Shiny uses similar storing conventions as R packages

--------------------------------------------------------------

## UI Functions 

**Functions are useful in reducing duplication**

CASE 1: converting multiple sliders to a function

```{r slider1, eval=FALSE}
ui <- fluidRow(
  sliderInput("alpha", "alpha", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("beta",  "beta",  min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("gamma", "gamma", min = 0, max = 1, value = 0.5, step = 0.1),
  sliderInput("delta", "delta", min = 0, max = 1, value = 0.5, step = 0.1)
)
```

TO: 

```{r slider2, eval=FALSE}
sliderInput01 <- function(id) {
  sliderInput(id, label = id, min = 0, max = 1, value = 0.5, step = 0.1)
}

ui <- fluidRow(
  sliderInput01("alpha"),
  sliderInput01("beta"),
  sliderInput01("gamma"),
  sliderInput01("delta")
)
```


Like functions in R analyses, functions in apps makes you app more readable and efficient if we need to change the behavior, we would only do it one place. 

CASE 2: customized `dateInput` with `...`

```{r date-input, eval=FALSE}
usWeekDateInput <- function(inputId, ...) {
  dateInput(inputId, ..., format = "dd M, yy", daysofweekdisabled = c(0, 6))
}
```

Just a reminder on `dateInput`:

```
dateInput(
  inputId,
  label,
  value = NULL,
  min = NULL,
  max = NULL,
  format = "yyyy-mm-dd",
  startview = "month",
  weekstart = 0,
  language = "en",
  width = NULL,
  autoclose = TRUE,
  datesdisabled = NULL,
  daysofweekdisabled = NULL
)
```

CASE 3: radio buttons to make 

```{r}
iconRadioButtons <- function(inputId, label, choices, selected = NULL) {
  names <- lapply(choices, icon)
  values <- if (is.null(names(choices))) names(choices) else choices
  radioButtons(inputId,
               label = label,
               choiceNames = names, choiceValues = values, selected = selected
  )
}
```


### Functional Programming

+ Using functions like `map()` may help reduce code further

```{r map-slider, eval=FALSE}
library(purrr)
# pass the variables for the slider to the built function sliderInput01
vars <- c("alpha", "beta", "gamma", "delta")
# returns a list of sliders
sliders <- map(vars, sliderInput01)
# fluidRow unpacks the list to become children of the container
ui <- fluidRow(sliders)
```


### UI as data

**Turning a UI structure into a data structure to have more varitey in inputs**

```{r ui-data, eval=FALSE}
vars <- tibble::tribble(
  ~ id,   ~ min, ~ max,
  "alpha",     0,     1,
  "beta",      0,    10,
  "gamma",    -1,     1,
  "delta",     0,     1,
)

# function where arg names match the col names

mySliderInput <- function(id, label = id, min = 0, max = 1) {
  sliderInput(id, label, min = min, max = max, value = 0.5, step = 0.1)
}

# pmap to call mySliderInput over vars

sliders <- pmap(vars, mySliderInput)
```

---------------------------------------

## Server Functions

+ Long reactives should be put into a non-reactive functions outside the server
  + easier to debug
  + easier to tell what inputs were passed to a function vs a reactive expression


CASE 1: Reading uploaded data

```{r data, eval=FALSE}

server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    
    ext <- tools::file_ext(input$file$name)
    switch(ext,
      csv = vroom::vroom(input$file$datapath, delim = ","),
      tsv = vroom::vroom(input$file$datapath, delim = "\t"),
      validate("Invalid file; Please upload a .csv or .tsv file")
    )
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}

```

To a function that extracts:

```{r data-extract, eval=FALSE}

# IN ITS OWN FILE
load_file <- function(name, path) {
  ext <- tools::file_ext(name)
  switch(ext,
    csv = vroom::vroom(path, delim = ","),
    tsv = vroom::vroom(path, delim = "\t"),
    # validate works similarly to stop() outside of shiny
    validate("Invalid file; Please upload a .csv or .tsv file")
  )
}
#----------------------------------------------
server <- function(input, output, session) {
  data <- reactive({
    req(input$file)
    load_file(input$file$name, input$file$datapath)
  })
  
  output$head <- renderTable({
    head(data(), input$n)
  })
}
```

Note: generally better to keep reactive and non-reactive parts of app as separate as possible

### Internal functions


If the function needs to use input, output or session, it makes more sense to write the function directly in the server

```{r internal-func, eval=FALSE}
server <- function(input, output, session) {
  switch_page <- function(i) {
    updateTabsetPanel(input = "wizard", selected = paste0("page_", i))
  }
  
  observeEvent(input$page_12, switch_page(2))
  observeEvent(input$page_21, switch_page(1))
  observeEvent(input$page_23, switch_page(3))
  observeEvent(input$page_32, switch_page(2))
}
```


-------------------------------------------


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/g927N1YRjAA")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:02:16	Federica Gazzelloni:	hello
00:02:32	russ:	Hi everyone
00:06:20	russ:	A couple of the regulars apologised that they couldn't make it this week
00:30:10	Federica:	https://shiny.rstudio.com/reference/shiny/1.6.0/radioButtons.html
00:34:44	Federica:	2.2.5 Limited choices
There are two different approaches to allow the user to choose from a prespecified set of options: selectInput() and radioButtons().
```
</details>

### Cohort 2

`r knitr::include_url("https://www.youtube.com/embed/aQHyWOEvJAI")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:27:24	Kevin Gilds:	https://www.rostrum.blog/2021/04/14/gha-readme/
00:27:42	Kevin Gilds:	Update README with Github actions
00:51:26	collinberke:	https://www.youtube.com/watch?v=BEyiFVWaTtc&list=PL3x6DOfs2NGjtn1_4BSX99R5wrLjK7XvY&index=19
00:51:35	collinberke:	https://www.youtube.com/watch?v=tWEa6JwlJ3c&list=PL3x6DOfs2NGjtn1_4BSX99R5wrLjK7XvY&index=21
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/5YmZiyVgUeg")`

### Cohort 3

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary>Meeting chat log</summary>
```
LOG
```
</details>
